Some lessons after the formalization of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-05-29 Independence_CH/proof_assists_isabelle.tex
First version of intro for section 2.

Proof assistants provide diverse aids for the task
of formalizing a piece of mathematics. They are usually implemented
using a typed programming language; rigor is enforced by defining a
type of ``theorems,'' whose members can only be constructed using
operations stipulated in a small \emph{kernel} which encodes the
underlying foundation of the assistant. Further developments interact
with the type of theorems only through the kernel, and thus the latter
is the only ``trusted'' part of the assistant's code.

Several of the more established assistants (HOL Light, Coq, Isabelle)
are programmed in some variant of the ML language (which was
designed for this purpose); the newer Lean, on the other
hand, was originally conceived as a standalone functional programming
language with all the features of a standard assistant.

In the case of Isabelle, Standard ML is the first of the four layers
in which we work in this assistant. Both the kernel and automation
of proofs is coded in ML, sometimes as a substitute for induction on
formulas, as the next section explains.

>>>>

The core of Isabelle, as well as LCF, is based on higher-order logic;
another heritage of LCF is the possibility for the user to define
\emph{tactics} to encapsulate common patterns used to solve goals. In
fact, this is the origin of the ML family of languages: a
Meta-Language for programming tactics. In contrast with LCF, Isabelle
was designed~\cite{Paulson1990} as a generic theorem for several
logics. Isabelle still offers a programmable interface (in ML) that
allows the user, for instance, to define a tactic solving a statement
about a natural number $n$ by doing cases on whether $n=0$ or
$n\neq 0$.  Moreover one can define functions that manipulate terms
(in the usual sense) or even formulas.

In contrast with LCF, some of the more prominent ITPs of today are
based on some type theory. Both Coq~\cite{coq} and
lean~\cite{DBLP:conf/cade/Moura021} are based on the Calculus of
Inductive Constructions
\cite{DBLP:conf/mfps/PfenningP89,DBLP:journals/iandc/CoquandH88},
while Agda~\cite{agda} is an extension of Martin-Löf type
theory~\cite{DBLP:books/daglib/0000395}. The oldest ITP still
still used today is Mizar~\cite{mizar} is far away both in terms
of foundations and architecture; it inspired, though, the Isar
dialect used in nowadays Isabelle.

% Proof assistants provide diverse aids for the task
% of formalizing a piece of mathematics. They are usually implemented
% using a typed programming language; rigor is enforced by defining a
% type of ``theorems,'' whose members can only be constructed using
% operations stipulated in a small \emph{kernel} which encodes the
% underlying foundation of the assistant. Further developments interact
% with the type of theorems only through the kernel, and thus the latter
% is the only ``trusted'' part of the assistant's code.

% Several of the more established assistants (HOL Light, Coq, Isabelle)
% are programmed in some variant of the ML language (which was
% designed for this purpose); the newer Lean, on the other
% hand, was originally conceived as a standalone functional programming
% language with all the features of a standard assistant.

% In the case of Isabelle, Standard ML is the first of the four layers
% in which we work in this assistant. Both the kernel and automation
% of proofs is coded in ML, sometimes as a substitute for induction on
% formulas, as the next section explains.
